---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  word_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, error=FALSE, collapse=TRUE)
```
# Do right handed MLB players hit more singles on average than left handed players? An analysis
## Introduction
Baseball is an asymetric game. Since players must reach each base in a counter clockwise order, it is assumed right handed hitters must have an advantage in hitting singles since their natural range is the right field and that field is the farthest away from the first base. This is what we looked to find out in this analysis, whether right handed MLB hitters hit more singles on average than left handed players. To get there, we selected the data from the [Lahman Baseball Databank](http://www.seanlahman.com/baseball-archive/statistics/), to which we cleaned the data and summarized the average number of singles per at bat for left handed and right handed players, and then did an hypothesis testing to see if the difference is statistically significative. The conclusions were that it was the opposite, that left handed players hit more singles on average than right handed players, with statistical significance. The causes of this difference are so far unknown and will be open for further analysis.

## Analysis
### Data
The data collected consisted from the archives of the Lahman Baseball Databank, which contains complete batting and pitching statistics from 1871 to 2017, plus fielding statistics, standings, team stats, managerial records, post-season data, and more, from every player and team from the listed time frame. More specifically, two data subsets were used:

 1. The *Batting* subset, which gave us the batting statistics per *playerID*, per season, since 1871.
 2. The *People* subset, which gave us the players infos from each *playerID*, including the needed batting position.

### Methods
The methods consisted of the initial importing and cleaning which was divided in these steps:

 1. Importing our needed libraries: dplyr and readr
```{r, results='markup', message=FALSE, error=FALSE, warning=FALSE}
# Importing the libraries
library(dplyr)
library(readr)
```

 2. Importing our two subsets: Batting.csv and People.csv
```{r, results='markup', message=FALSE, error=FALSE, warning=FALSE}
# Importing the two subsets
Batting <- read_csv("C:/Users/Guillaume/Downloads/baseballdatabank-master_2018-03-28/baseballdatabank-master/core/Batting.csv")
People <- read_csv("C:/Users/Guillaume/Downloads/baseballdatabank-master_2018-03-28/baseballdatabank-master/core/People.csv")
```
 
 3. Joining our two subsets: Frap
```{r, results='markup', message=FALSE, error=FALSE, warning=FALSE}
# Joining the two subsets
Frap <- inner_join(Batting, People, by = "playerID")
```
 
 4. Arrange our Frap subset to get our needed data frame Frappeurs, which will include our playersID, his batting position and his single's average over his MLB career.
```{r, results='markup', message=FALSE, error=FALSE, warning=FALSE}
# Creating the Frappeurs dataframe
Frappeurs <- Frap %>%
  select(playerID, yearID, AB, H, "2B", "3B", HR, bats) %>%
  mutate(simples = H - `2B` - `3B` - HR) %>%
  select(playerID, yearID, bats, AB, simples) %>%
  filter(bats %in% c("R", "L") & AB >= 100) %>%
  mutate(propsimple = simples/AB) %>%
  select(playerID, yearID, bats, propsimple) %>%
  group_by(playerID, bats) %>%
  summarize(avgpropsimple = mean(propsimple))
```

Then, we verified the suppositions in order to make sure we could operate an hypothesis testing. These suppositions were the:

1. Normality (through an Anderson-Darling test)
2. Homoscedasticity (through a boxplot)
 
These steps are listed in this code:
```{r, results='hold', message=FALSE, error=FALSE, warning=FALSE}
# Verification of the suppositions
# Creation of the residual subsets for the Anderson-Darling test
droitier <- Frappeurs[Frappeurs$bats == "R", ]
droitier.res <- droitier$avgpropsimple - mean(droitier$avgpropsimple)
gaucher <- Frappeurs[Frappeurs$bats == "L", ]
gaucher.res <- gaucher$avgpropsimple - mean(gaucher$avgpropsimple)
Frappeurs$Res <- c(droitier.res, gaucher.res)

# Verification of normality through an Anderson-Darling test
library("nortest")
ad.test(Frappeurs$Res)
```

```{r, results='hold', message=FALSE, error=FALSE, warning=FALSE}
# Graphic display of the test
options(jupyter.plot_mimetypes = 'image/png')
par(mfrow = c(1, 2))
qqnorm(Frappeurs$Res, ylab = "Observed quantiles",
       xlab = "Theoric quantiles",
       main = "Quantile-quantile plot")
qqline(Frappeurs$Res)
text(x = -2, y = 14.5, labels = "a", cex = 1.2)

# Verification of homoscedasticity through a boxplot
boxplot(Res ~ bats, data = Frappeurs)
text(x = 0.5, y = 14.5, labels = "b", cex = 1.2)
```

The subsets are normalitated and homoscedasticitated and thus we could conduct the statistical student test to verify the starting premise.

```{r, results='hold', message=FALSE, error=FALSE, warning=FALSE}
# Conducting Student Test
t.out <- t.test(avgpropsimple ~ bats, data = Frappeurs, var.equal = TRUE,
                alternative = "less")
t.out
```

### Interpretation

The results of the Student Test shows that right handed MLB hitters *do not* hit more singles on average, and that the differences are statistically significant.


### Results

The results illustrate that not only right handed MLB hitters do not hit more singles on average, but they actually hit less than left handed hitters, with the difference being statistically significant. Here are plotted the results, with 95% confidence intervals:

```{r, results='hold', message=FALSE, error=FALSE, warning=FALSE}
# Group averages
moy.droitier <- mean(droitier$avgpropsimple)
moy.gaucher <- mean(gaucher$avgpropsimple)
# Group SD
sd.droitier <- sd(droitier$avgpropsimple)
sd.gaucher <- sd(gaucher$avgpropsimple)
# Group Sizes
n.droitier <- nrow(droitier)
n.gaucher <- nrow(gaucher)
# Group Standard Errors
SE.droitier <- sd.droitier/sqrt(n.droitier)
SE.gaucher <- sd.gaucher/sqrt(n.gaucher)
# Confidence Intervals (set at 95%)
IC.inf95.droitier <- moy.droitier - qt(p = 0.025, df = n.droitier - 1) * SE.droitier
IC.sup95.droitier <- moy.droitier + qt(p = 0.025, df = n.droitier - 1) * SE.droitier
IC.inf95.gaucher <- moy.gaucher - qt(p = 0.025, df = n.gaucher - 1) * SE.gaucher
IC.sup95.gaucher <- moy.gaucher + qt(p = 0.025, df = n.gaucher - 1) * SE.gaucher
# Plot
plot(y = 0, x = 0, xlab = "Type of hitter",
     ylab = "Proportion of singles per at bat",
     main = "Averages ± CI at 95%",
     ylim = range(c(IC.inf95.droitier, IC.inf95.gaucher,
                    IC.sup95.droitier, IC.sup95.gaucher)),
     xlim = c(0, 1),
     xaxt = "n")
# X-Axis
axis(side = 1, at = c(0.3, 0.75), labels = c("right handed", "left handed"))
# Scatters
points(y = c(moy.droitier, moy.gaucher), x = c(0.3, 0.75))
# Confidence Intervals
arrows(x0 = c(0.3, 0.75), x1 = c(0.3, 0.75),
       y0 = c(IC.inf95.droitier, IC.inf95.gaucher),
       y1 = c(IC.sup95.droitier, IC.sup95.gaucher),
       length = 0.05, code = 3, angle = 90)
```

## Conclusion
In conclusion, after analysing 6048 MLB hitters from 1871 to 2017, we found that right handed hitters do not hit more singles on average than left handed hitters. Rather, the opposite holds, that left handed hitters hit more singles on average than right handed ones. Results were statistically significant.


## Appendix
Lahman Baseball Databank: http://www.seanlahman.com/baseball-archive/statistics/
Original R Script: https://github.com/Guillaume2509/MLBSingleAvg/blob/master/MLBSingleAvg.R

Guillaume Giroux, 2019